---
title: "Class05: Data Viz with GGPLOT"
author: "Parnaz Boroon (PID:A13557370)"
format: gfm
---

Today we are playing with plotting and graphing in R.

There are lots of ways to make cool figures in R.
There is "base" R graphics ('plot()', 'hist()', 'boxplot()', etc.)

There is also add-on packages, like **ggplot**.

```{r}
head(cars,3)
```
Let's plot this with "base" R:
```{r}
plot(cars)
```
```{r}
head(mtcars)
```

Let's plot 'mpg' vs. 'disp'
```{r}
plot(mtcars$mpg,mtcars$disp, pch=8)
```

```{r}
hist(mtcars$mpg)
```

## GGPLOT

The main function in the ggplot2 package is 'ggplot()'. First I need to install the **ggplot2** package. I can install any package with the function 'install.(packages)'. You don't want to keep downloading ggplot over and over again. *In order to not download the software package over and over again, type this code in the console.*

>**N.B.** I never want to run 'install.packages()' in my quarto source document girl!

```{r}
library(ggplot2)
ggplot(cars) +
  aes(x=speed, y=dist)+
  geom_point()
```

```{r}
ggplot(cars)+
  aes(speed)+
  geom_histogram()
```

Every ggplot needs at least 3 things:

- The **data** (given with 'ggplot(cars)')
- The **aesthetics** mapping (given with 'aes()')
- The **geometry** (given with geom_point)

>For simple canned graphs "base" R is nearly always faster

### Adding more layers

Let's add a line and a title, subtitle and caption as well as custom axis labels.


```{r}
library(ggplot2)
ggplot(cars) +
  aes(x=speed, y=dist)+
  geom_point() +
  geom_line ()
```

```{r}
library(ggplot2)
ggplot(cars) +
  aes(x=speed, y=dist)+
  geom_point() +
  geom_smooth ()
```

```{r}
library(ggplot2)
ggplot(cars) +
  aes(x=speed, y=dist)+
  geom_point() +
  geom_smooth (method="lm", se=FALSE)
```

```{r}
library(ggplot2)
ggplot(cars) +
  aes(x=speed, y=dist)+
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs (
    title= "Parnaz's Silly Plot",
    subtitle = "This is my practice plot.",
    x="Speed(MPH)",
    y="Distance (ft)")+
    theme_bw()
```

##Plot some expression data 

Read data file from online URL

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
> Q1. How many genes are in this wee dataset?

There are `r nrow(genes)` in this dataset.

> Q2. How many "up" regulated genes are there?

```{r}
sum(genes$State == "up")
```

These are `r sum(genes$State == "up")` genes.

```{r}
table(genes$State)
```


```{r}
ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State)+
  geom_point()+
  scale_colour_manual( values=c("blue","gray","red") )+
   labs (
    title= "Gene Expression Changes Upon Drug Treatment",
    x="Control(no drug)",
    y="Drug Treatment")
```

Silly example of adding labels

```{r}
library(ggrepel)

ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State,label=Gene)+
  geom_point()+
  scale_colour_manual( values=c("blue","gray","red") )+
   labs (
    title= "Gene Expression Changes Upon Drug Treatment",
    x="Control(no drug)",
    y="Drug Treatment")+
  geom_text_repel(max.overlaps = 100)
```

## Going Further

Playing with some different layers and the gapmider dataset...

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder)
```
```{r}
tail(gapminder)
```
```{r}
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, col=continent) +
  geom_point()+
  facet_wrap(~continent)
```
Insert fun plot here:
```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp,
      col = continent,
      size = pop) +                 # bubble size reflects population
  geom_point(alpha = 0.7) +
  facet_wrap(~continent) +
  scale_size(range = c(1, 8)) +      # keep bubbles readable
  labs(size = "Population (log‑scaled)") +
  theme_minimal()
```


