---
title: "Class10"
author: "Parnaz Boroon PID:A13557370"
format: gfm
---
```{r}
candy_file <- "candy-data.csv"

candy <- read.csv(candy_file, row.names = 1)
head(candy)
```
## Q1. How many different candy types are in this dataset?
```{r}
nrow(candy)
```
There are 85 candy types. 

## Q2. How many fruity candy types are in the dataset?
```{r}
sum(candy$fruity == 1)
colnames(candy)
```
There are 38 fruit candy types.

```{r}
candy["Twix", ]$winpercent
```

## Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
```{r}
candy["Air Heads", "winpercent"]
grep("Air", rownames(candy), value = TRUE)
```
Air Heads and 52.34146.


## Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy["Kit Kat", "winpercent"]
grep("Kit Kat", rownames(candy), value = TRUE)
```

## Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
```{r}
candy["Tootsie Roll Snack Bars", "winpercent"]
grep("Tootsie Roll Snack Bars", rownames(candy), value = TRUE)
```
```{r, message-FALSE}
library(dplyr)

candy |> nrow()
```

```{r}
nrow(candy)
```

```{r}
round( mean(candy$winpercent),4)
```

```{r}
win <- candy$winpercent
win.mean <- mean(win)
round(win.mean)

```

```{r}
candy %>%select(winpercent)
candy |>
  select(winpercent)
```
```{r}
library("skimr")
skim(candy)
```
## Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
The winpercent row looks to be on a different scale.

## Q7. What do you think a zero and one represent for the candy$chocolate column?
0 and 1 means either candy or chocolate. 1 is chocolate

## Q8. Plot a histogram of winpercent values
```{r}
library(ggplot2)

ggplot(candy)+
  aes(winpercent)+
  geom_histogram()
```


## Q9. Is the distribution of winpercent values symmetrical?
```{r}
ggplot(candy)+
  aes(winpercent)+
  geom_density()
```
## Q10. Is the center of the distribution above or below 50%?
```{r}
mean(candy$winpercent)
```
```{r}
summary(candy$winpercent)
```

## Q11. On average is chocolate candy higher or lower ranked than fruit candy?
```{r}
# 1. Find all chocolate candy in the dataset.
choc.inds <- as.logical(candy$chocolate)
choc.candy <- candy[choc.inds,]
choc.candy
# 2. Extract their `winpercent` values
choc.win <- choc.candy$winpercent
choc.win
# 3. Find the mean of these values
mean(choc.win)
# 4-6. Do the same for fruity candy.
fruits.inds <- as.logical(candy$fruity)
fruits.candy <- candy [fruits.inds,]
fruits.candy

fruits.win <- fruits.candy$winpercent
fruits.win

mean(fruits.win)

# 7. Which mean value is higher?
#Chocolate candy mean is higher.
```

## Q12. Is this difference statistically significant?
yes, significantly different
```{r}
t.test(choc.win, fruits.win)
```
## Q13. What are the five least liked candy types in this set?
```{r}
x<-c(10,2,5,1)
order(x)
```
## Q14. What are the top 5 all time favorite candy types out of this set?
```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind,],5)
```

## Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```
## Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),-winpercent)) +
  geom_col()
```

Add some color
```{r}
my_cols <- rep("black",nrow(candy))
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
my_cols

ggplot(candy) + 
  aes(x=winpercent,
      y=reorder(rownames(candy),-winpercent)) +
  geom_col(fill=my_cols)
```
## Q17: What is the worst ranked chocolate candy?
Sixlets 
## Q18: What is the best ranked fruity candy?
Starburst


```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols)
```

## Q19: Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
tail( candy[ord,c(11,12)], n=5 )

```

## Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```



```{r}
cor(candy)
```

```{r}
cij<- cor(candy)
```

```{r}
library(corrplot)
corrplot(cij)
```
## Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
Fruity and chocolate

## Q23. Similarly, what two variables are most positively correlated?
chocolate and winpercent

The main function in base R for this `prcomp()`and we want to set `scale=TRUE` here:
```{r}
pca <- prcomp(candy, scale. = TRUE)
summary(pca)
```

```{r}
ggplot(pca$x)+
  aes(PC1,PC2)+
  geom_point(col=my_cols)
```
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
```


```{r}
ggplot(pca$rotation)+
  aes(PC1, rownames(pca$rotation))+
  geom_col()
```

## Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
Pluribus, hard, and fruity are all in the positive direction which does make sense to me because they all have similar characteristics.
